Token-level author diarization using clustering of stylistic contexts

### 2017-06-01 paper/paper.tex
Chapter 4 finished.

% details for each component
\paragraph{Tokenization.} As a preprocessing step, we tokenized each document $D$ using NLTK\footnote{\url{http://www.nltk.org}} to obtain a set of tokens $T_D$. We also performed part-of-speech tagging with the same tool, to speed up basic feature extraction which we describe below. We didn't use other preprocessing techniques such as lemmatization, stemming and stop word removal because they would take away a lot of stylometric data from text \cite{stamatatos-2009a}. The final output from the tokenization step was a finite sequence $\{(t_i, o_i, l(t_i), \mathit{POS}_i)\}_{i=1}^n$ where $o_i$ is the offset of token $t_i$, $l(t_i)$ its length in characters and $\mathit{POS}_i$ its POS tag.

\paragraph{Basic feature extraction.} 
We defined the stylistic context of a token $t_i$ as a set of tokens $\{t_k\}_{k=i-cs}^{i-1} \cup \{t_k\}_{k=i+1}^{i+cs}$ where $cs$ is a context size. From each context we extracted the most used stylometric features which we found in previous work.
\begin{itemize}
	\item \textit{Character tri-grams}. Frequencies of \textit{n}-grams on character level have been very useful in quantifying the writing style \citep{stamatatos-2009b}. They are able to capture lexical and contextual information, use of punctuation and errors which can be an author's  "fingerprint". This feature is also tolerant to noise. Based on work by \citet{stamatatos-2009a} and \citet{rahman-2015}, we choose $n=3$. Maximal dimension of this feature vector was set to 200.
	\item \textit{Stop words}. According to \citet{stamatatos-2009a}, sometimes also called \textit{function words}, these are the most common used topic-independent words in text, such as articles, prepositions, pronouns and others. They are used unconciously and found to be one of the most discriminative features in authorship attribution since they represent pure stylistic choices of authors' \citep{burrows-1987,argamon-2005}. We used frequencies of 156 english stop words available in \texttt{NLTK}.
	\item \textit{Special characters}. We used counts of all character sequences which satisfied a regular expression defined as \verb/[^\\w]{1,4}/. Although character $n$-grams can catch the use of those character sequences, we wanted to have a distinct feature for that purpose since \citet{koppel-2009} mentioned that authors can have different punctuation habbits.
	\item \textit{POS tag counts}. This is syntactic feature which \citet{koppel-2009} and \citet{stamatatos-2009a} also identified as a discriminative one in authorship analysis and it was used by \citet{kuznetsov-2016}. We used all 12 tags from the universal tagset.
	\item \textit{Average token length}. Used by \citet{kuznetsov-2016}, \citet{sittar-2016},  \citet{brooke-2012} and \citet{stein-2011}. \citet{koppel-2009} characterized this feature as a complexity measure.
	\item \textit{Bag of Words}. Bag of words text representation more captures content, rather than style \citep{stamatatos-2009a}. We included this feature because it boosted performance of our initial testing, even without words been previously stemmed or lemmatized (stop words were excluded). Counts of maximum of 100 unigrams were used.
	\item \textit{Type-token ratio}. We wanted to use that feature to measure the vocabulary richness in a token's context, but after we finished with performance evaluations we realized that there was a bug in implementation and that feature did not contribute to overal results. The feature should be calculated as the ratio of vocabulary size and total number of tokens of the text \citep{stamatatos-2009a}.
\end{itemize}

We implemented two variants of our final model. First is with only a basic feature extractor $f_\mathrm{b}$, whether in second the feature transformation $f_\mathrm{t}$ was used. For the first variant, where only a basic feature extractor $f_\mathrm{b}$ was used, we extracted features for each document separately. In the variant with both basic extractor $f_\mathrm{b}$ and transformation $f_\mathrm{t}$, all documents were concatenated to a single one from which basic and transformed feature vectors were obtained. In both approaches basic features were extracted from a sliding window which moved from token to token and included a context of each token. In the rest of the paper we refer to the first variant as model A, and to the second as model B.

\paragraph{Feature transformation.} We develop two variants of our model which are mainly differentiated by the application of a trainable feature transformation requiring fixed basic feature space as opposed to variable feature space with features that can be more document-specific.

Let $(\vect{b}_i)_{i=1}^n = f_\mathrm{b}(D)$ be a sequence of basic feature vectors and $(\vect{b}_i')_{i=1}^n$ the corresponding sequence of potentially preprocessed basic feature vectors with elements from $\mathbb{R}^{n_{b}'}$. Let $(a_i)_{i=1}^n$ be the sequence of true author labels with elements from $\{1..c\}$. We want to maximize the \emph{clusterability} of the feature vectors obtained by the feature transformation $T:\mathbb{R}^{n_b'}\rightarrow\mathbb{R}^{n_t}$ with trainable parameters $\vect{\theta}_T$, i.e. we want to maximize segregation and compactness of groups of transformed feature vectors grouped by their target author label. Let $(\vect{t}_i)_{i=1}^n = (T(\boldsymbol{b}_i'))_{i=1}^n = f_\mathrm{t}((\vect{b}_i)_{i=1}^n)$ be the sequence of transformed feature vectors with elements from $\mathbb{R}^{n_t}$. We define the following loss in order to optimize the transformation $T$:
\begin{equation}
	L = \alpha L_\mathrm{c} + (1-\alpha)L_\mathrm{s}
\end{equation}
with $\alpha$ chosen to be $0.5$. Here $L_\mathrm{c}$ is the \emph{compactness loss} and $L_\mathrm{s}$ the \emph{segregation loss}. $L_\mathrm{c}$ is proportional to the average variance of groups, and $L_\mathrm{s}$ penalizes centroids being too close to each other. Let $N_a$ represent the number of tokens associated with author $a$, i.e. $N_a = \sum_{i=1}^n[a_i=a]$, where, for $\varphi$ being a logical proposition $[\varphi]$ evaluates to either $1$ if $\varphi$ is true or $0$ otherwise.
%, the following notation\footnote{It is called \emph{the Iverson bracket}.} is introduced:
%\begin{equation}
%[\varphi] = \begin{cases}
%	1, & \varphi\equiv\top, \\
%	0, & \varphi\equiv\bot.
%\end{cases}
%\end{equation}
Let $\vect{\mu}_a$ represent the current centroid of the transformed feature vectors associated with the author label $a$:
\begin{equation}
\vect{\mu}_a = \frac{1}{N_a}\sum_{i=1}^n T(\vect{b}_i')[a_i=a]
\end{equation}	
We define the components of the loss: 
\begin{align}
	L_\mathrm{c} &= \sum_{a=0}^{c}\frac{1}{N_a} 
		\sum_{i=1}^{n} \| T(\vect{b}_i')-\vect{\mu}_{a}\|^2 [a_i=a], \\
	L_\mathrm{s} &= \frac{2}{c}\sum_{a=0}^{c}\sum_{b=a+1}^{c} \|\vect{\mu}_a-\vect{\mu}_b\|^{-2}.
\end{align}
The \emph{compactness loss} $L_\mathrm{c}$ is a weighted sum of within-group variances. The \emph{segregation loss}  $L_\mathrm{s}$ is proportional to the sum of magnitudes (not magnitude of the sum) of forces between $c$ equally charged particles each located at one of the centroids. By minimizing the average loss (the error) across all documents in the training dataset with respect to the parameters of the transformation $\vect{\theta}_T$ we hope to benefit the clustering algorithm. Instead of squared $L^2$ distance, some other distance measure may be used. Also, the way of combining the two losses was somewhat arbitrarily chosen as well as the segregation loss. For the transformation $T$ we have tried a nonlinear transformations represented as neural networks with one or more hidden layers. We didn't explore it much because it was slower and didn't give as good results as a linear transformation or an elementwise product with a weight vector:
\begin{align}
	T_1(\vect{x}) &= \matr{W}\vect{x}, \quad \vect{\theta}_{T_1} = \matr{W}, \\
	T_2(\vect{x}) &= \vect{w}\odot\vect{x}, \quad \vect{\theta}_{T_2} = \vect{w}.
\end{align}
Before applying the transformations, basic feature vectors are preprocessed by concatanating each with the vector of its squared elements:
\begin{equation}
	\vect{b_i'} = (\vect{b_i}^\transpose, (\vect{b_i}\odot\vect{b_i})^\transpose)^\transpose, \quad i\in \{1..n\}.
\end{equation}
The transformation is implemented using TensorFlow\footnote{\url{https://www.tensorflow.org}}. It is trained on-line (one document per optimization step) using RMSProp \citep{tielman-2012} with default parameters and learning rate $5\times 10^{-4}$.

@@

The final step in our approach was clustering of obtained feature vectors which represented stylistic contexts of tokens. The number of clusters was equal to the number of authors in tasks \emph{a} and \emph{b}, but in task \emph{c} it was unknown. We tested various clustering algorithms available in \texttt{scikit-learn}\footnote{\url{http://scikit-learn.org}} Python toolkit by \citet{scikit-learn} in all tasks. In the first variant of our approach for tasks  \emph{a} and \emph{b} we propose hierarchical agglomerative clustering, and for task \emph{c} DBSCAN algorithm which estimates the number of clusters automatically. In the second variant we used k-means for clustering in all tasks, and tried our modification of k-means, which we called Auto-k-means in task \emph{c}. The reason for choosing those algorithms is described in section \ref{sec:experimental-results}.

>>>>

% details for each component
\paragraph{Tokenization.} As a preprocessing step, we tokenized each document $D$ using \texttt{NLTK}\footnote{\url{http://www.nltk.org}} by \citet{bird-2009} to obtain a set of tokens $T_D$. We also performed part-of-speech tagging with the same tool, to speed up basic feature extraction which we describe below. We didn't use other preprocessing techniques such as lemmatization, stemming and stop word removal because they would take away a lot of stylometric data from text \cite{stamatatos-2009a}. The final output from the tokenization step was a finite sequence $\{(t_i, o_i, l(t_i), \mathit{POS}_i)\}_{i=1}^n$ where $o_i$ is the offset of token $t_i$, $l(t_i)$ its length in characters and $\mathit{POS}_i$ its POS tag.

\paragraph{Basic feature extraction.} 
We defined the stylistic context of a token $t_i$ as a set of tokens $\{t_k\}_{k=i-cs}^{i-1} \cup \{t_k\}_{k=i+1}^{i+cs}$ where $cs$ is a context size. From each context we extracted the most used stylometric features which we found in previous work.
\begin{itemize}
	\item \textit{Character tri-grams}. Frequencies of \textit{n}-grams on character level have been very useful in quantifying the writing style \citep{stamatatos-2009b}. They are able to capture lexical and contextual information, use of punctuation and errors which can be an author's  "fingerprint". This feature is also tolerant to noise. Based on work by \citet{stamatatos-2009a} and \citet{rahman-2015}, we choose $n=3$. Maximal dimension of this feature vector was set to 200.
	\item \textit{Stop words}. According to \citet{stamatatos-2009a}, sometimes also called \textit{function words}, these are the most common used topic-independent words in text, such as articles, prepositions, pronouns and others. They are used unconciously and found to be one of the most discriminative features in authorship attribution since they represent pure stylistic choices of authors' \citep{burrows-1987,argamon-2005}. We used frequencies of 156 english stop words available in \texttt{NLTK}.
	\item \textit{Special characters}. We used counts of all character sequences which satisfied a regular expression defined as \verb/[^\\w]{1,4}/. Although character $n$-grams can catch the use of those character sequences, we wanted to have a distinct feature for that purpose since \citet{koppel-2009} mentioned that authors can have different punctuation habbits.
	\item \textit{POS tag counts}. This is syntactic feature which \citet{koppel-2009} and \citet{stamatatos-2009a} also identified as a discriminative one in authorship analysis and it was used by \citet{kuznetsov-2016}. We used all 12 tags from the universal tagset.
	\item \textit{Average token length}. Used by \citet{kuznetsov-2016}, \citet{sittar-2016},  \citet{brooke-2012} and \citet{stein-2011}. \citet{koppel-2009} characterized this feature as a complexity measure.
	\item \textit{Bag of Words}. Bag of words text representation more captures content, rather than style \citep{stamatatos-2009a}. We included this feature because it boosted performance of our initial testing, even without words been previously stemmed or lemmatized (stop words were excluded). Counts of maximum of 100 unigrams were used.
	\item \textit{Type-token ratio}. We wanted to use that feature to measure the vocabulary richness in a token's context, but after we finished with performance evaluations we realized that there was a bug in implementation and its impact on final results was unknown to us. The feature should be calculated as the ratio of vocabulary size and total number of tokens of the text \citep{stamatatos-2009a}.
\end{itemize}

\paragraph{Feature transformation.}

 We develop two variants of our model which are mainly differentiated by the application of a trainable feature transformation requiring fixed basic feature space as opposed to variable feature space with features that can be more document-specific.

 First variant uses only a basic feature extractor $f_\mathrm{b}$, whether in second the feature transformation $f_\mathrm{t}$ is used. In the first variant we extract features for each document separately and scale features to zero mean and unit variance. In the variant with both basic extractor $f_\mathrm{b}$ and transformation $f_\mathrm{t}$, all documents are concatenated to a single one from which basic and transformed feature vectors are obtained, but without standardization. In both approaches basic features are extracted from a sliding window which moves from token to token and includes a context of each token. In the rest of the paper we refer to the first variant as model A, and to the second as model B.

In model B, let $(\vect{b}_i)_{i=1}^n = f_\mathrm{b}(D)$ be a sequence of basic feature vectors and $(\vect{b}_i')_{i=1}^n$ the corresponding sequence of potentially preprocessed basic feature vectors with elements from $\mathbb{R}^{n_{b}'}$. Let $(a_i)_{i=1}^n$ be the sequence of true author labels with elements from $\{1..c\}$. We want to maximize the \emph{clusterability} of the feature vectors obtained by the feature transformation $T:\mathbb{R}^{n_b'}\rightarrow\mathbb{R}^{n_t}$ with trainable parameters $\vect{\theta}_T$, i.e. we want to maximize segregation and compactness of groups of transformed feature vectors grouped by their target author label. Let $(\vect{t}_i)_{i=1}^n = (T(\boldsymbol{b}_i'))_{i=1}^n = f_\mathrm{t}((\vect{b}_i)_{i=1}^n)$ be the sequence of transformed feature vectors with elements from $\mathbb{R}^{n_t}$. We define the following loss in order to optimize the transformation $T$:
\begin{equation}
	L = \alpha L_\mathrm{c} + (1-\alpha)L_\mathrm{s}
\end{equation}
with $\alpha$ chosen to be $0.5$. Here $L_\mathrm{c}$ is the \emph{compactness loss} and $L_\mathrm{s}$ the \emph{segregation loss}. $L_\mathrm{c}$ is proportional to the average variance of groups, and $L_\mathrm{s}$ penalizes centroids being too close to each other. Let $N_a$ represent the number of tokens associated with author $a$, i.e. $N_a = \sum_{i=1}^n[a_i=a]$, where, for $\varphi$ being a logical proposition $[\varphi]$ evaluates to either $1$ if $\varphi$ is true or $0$ otherwise.
%, the following notation\footnote{It is called \emph{the Iverson bracket}.} is introduced:
%\begin{equation}
%[\varphi] = \begin{cases}
%	1, & \varphi\equiv\top, \\
%	0, & \varphi\equiv\bot.
%\end{cases}
%\end{equation}
Let $\vect{\mu}_a$ represent the current centroid of the transformed feature vectors associated with the author label $a$:
\begin{equation}
\vect{\mu}_a = \frac{1}{N_a}\sum_{i=1}^n T(\vect{b}_i')[a_i=a]
\end{equation}	
We define the components of the loss: 
\begin{align}
	L_\mathrm{c} &= \sum_{a=0}^{c}\frac{1}{N_a} 
		\sum_{i=1}^{n} \| T(\vect{b}_i')-\vect{\mu}_{a}\|^2 [a_i=a], \\
	L_\mathrm{s} &= \frac{2}{c}\sum_{a=0}^{c}\sum_{b=a+1}^{c} \|\vect{\mu}_a-\vect{\mu}_b\|^{-2}.
\end{align}
The \emph{compactness loss} $L_\mathrm{c}$ is a weighted sum of within-group variances. The \emph{segregation loss}  $L_\mathrm{s}$ is proportional to the sum of magnitudes (not magnitude of the sum) of forces between $c$ equally charged particles each located at one of the centroids. By minimizing the average loss (the error) across all documents in the training dataset with respect to the parameters of the transformation $\vect{\theta}_T$ we hope to benefit the clustering algorithm. Instead of squared $L^2$ distance, some other distance measure may be used. Also, the way of combining the two losses was somewhat arbitrarily chosen as well as the segregation loss. For the transformation $T$ we have tried a nonlinear transformations represented as neural networks with one or more hidden layers. We didn't explore it much because it was slower and didn't give as good results as a linear transformation or an elementwise product with a weight vector:
\begin{align}
	T_1(\vect{x}) &= \matr{W}\vect{x}, \quad \vect{\theta}_{T_1} = \matr{W}, \\
	T_2(\vect{x}) &= \vect{w}\odot\vect{x}, \quad \vect{\theta}_{T_2} = \vect{w}.
\end{align}
Before applying the transformations, basic feature vectors are preprocessed by concatanating each with the vector of its squared elements:
\begin{equation}
	\vect{b_i'} = (\vect{b_i}^\transpose, (\vect{b_i}\odot\vect{b_i})^\transpose)^\transpose, \quad i\in \{1..n\}.
\end{equation}
The transformation is implemented using TensorFlow\footnote{\url{https://www.tensorflow.org}}. It is trained on-line (one document per optimization step) using RMSProp \citep{tielman-2012} with default parameters and learning rate $5\times 10^{-4}$.

@@

The final step in our approach was clustering of obtained feature vectors which represented stylistic contexts of tokens. The number of clusters was equal to the number of authors in tasks \emph{a} and \emph{b}, but in task \emph{c} it was unknown. We tested various clustering algorithms available in \texttt{scikit-learn}\footnote{\url{http://scikit-learn.org}} Python toolkit by \citet{scikit-learn} in all tasks. In the first variant of our approach for tasks  \emph{a} and \emph{b} we propose hierarchical agglomerative clustering, and for task \emph{c} DBSCAN algorithm which estimates the number of clusters automatically. In the second variant we used k-means for clustering in all tasks, and tried our modification of k-means, which we called Auto-k-means in task \emph{c}. The reason for choosing those algorithms is described in section \ref{sec:experimental-results}
